---
title: "INFO 6105 Final Project"
subtitle: "Impact of Sleep Duration on Academic Performance (GPA)"
author: "Rushita Vachhani | 002034308 | Section 37198"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
set.seed(6105) # Reproducibility

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 5.5, fig.align = "center"
)

library(tidyverse)
library(broom)
library(car)
library(pROC)
library(caret)
library(here)
library(GGally)
library(corrplot)
library(patchwork)
library(viridis)
library(knitr)
library(scales)
library(purrr)
library(randomForest)
library(glmnet)
```

```{=html}
<style>
body { font-family: 'Palatino', 'Georgia', serif; font-size: 15px; }
h1, h2 { color: #C8102E; border-bottom: 2px solid #C8102E; padding-bottom: 4px; }
h3 { color: #333333; }
.fill-box {
  border: 2px solid #C8102E; padding: 12px; margin: 10px 0;
  background-color: #fdf5f5; border-radius: 5px;
}
.concept-box {
  border: 2px solid #2166AC; padding: 12px; margin: 10px 0;
  background-color: #f0f6fc; border-radius: 5px;
}
.result-box {
  border: 2px solid #22863a; padding: 12px; margin: 10px 0;
  background-color: #f0fff4; border-radius: 5px;
}
</style>
```

---

# Introduction

This project investigates whether **sleep duration impacts academic performance
(GPA)** using real student wearable + academic data from Carnegie Mellon
University's Statistics & Data Science Repository.

We apply multiple statistical methods - Multiple Linear Regression, ANOVA,
Logistic Regression, Nonparametric Validation, and a Machine Learning pipeline
- to rigorously evaluate this relationship while controlling for cumulative
GPA and course load.

::: concept-box
**Research Questions:**

1. Does sleep duration significantly affect term GPA after controlling for cumulative GPA and course load?
2. Does mean GPA differ across study groups (universities/cohorts)?
3. How does sleep influence the probability of achieving a high GPA (≥ 3.0)?
4. Are findings robust under nonparametric assumptions?
:::

---

# Data Preparation

```{r load-data}
df_raw <- read.csv(here("data/raw/cmu-sleep.csv"))

df <- df_raw %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ, study) %>%
  mutate(
    TotalSleepTime = TotalSleepTime / 60,           # minutes -> hours
    high_gpa       = factor(if_else(term_gpa >= 3.0, 1, 0), levels = c(0, 1)),
    study          = factor(study),
    sleep_cat      = factor(
      case_when(
        TotalSleepTime < 6  ~ "Low (<6h)",
        TotalSleepTime <= 8 ~ "Medium (6-8h)",
        TRUE                ~ "High (>8h)"
      ),
      levels = c("Low (<6h)", "Medium (6-8h)", "High (>8h)")
    )
  )

# Complete-case dataset for all models
df_model <- df %>%
  drop_na(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ, study, high_gpa)
```

## Dataset Overview

```{r missingness-table}
# Missingness summary (for report Section 2)
tibble(
  Variable    = c("term_gpa", "TotalSleepTime (hrs)", "cum_gpa",
                  "Zterm_units_ZofZ", "study (groups)"),
  N_complete  = c(sum(!is.na(df$term_gpa)), sum(!is.na(df$TotalSleepTime)),
                  sum(!is.na(df$cum_gpa)),
                  sum(!is.na(df$Zterm_units_ZofZ)),
                  sum(!is.na(df$study))),
  Missing     = c(sum(is.na(df$term_gpa)), sum(is.na(df$TotalSleepTime)),
                  sum(is.na(df$cum_gpa)), sum(is.na(df$Zterm_units_ZofZ)),
                  sum(is.na(df$study))),
  Pct_Missing = paste0(round(Missing / nrow(df) * 100, 1), "%")
) %>%
  kable(caption = "Table 1: Variable Completeness Summary")
```

```{r descriptive-stats}
# Descriptive statistics (for report Section 2)
df_model %>%
  summarise(
    across(c(term_gpa, TotalSleepTime, cum_gpa),
           list(Mean = ~round(mean(.), 2), SD = ~round(sd(.), 2),
                Min  = ~round(min(.), 2),  Max = ~round(max(.), 2)),
           .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"),
               names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  kable(caption = "Table 2: Descriptive Statistics (Complete Cases)")
```

::: fill-box
**Key Preparation Steps**

- Sleep time converted from minutes to hours for interpretability
- Binary outcome `high_gpa` created (1 = term GPA ≥ 3.0)
- Sleep categories: Low < 6h, Medium 6–8h, High > 8h
- Complete-case dataset used for all models: **`r nrow(df_model)` observations**
:::

---

# Exploratory Data Analysis

## Outcome & Predictor Distributions

```{r eda-distributions, fig.cap="Figure 1: Distribution of term GPA (left) and sleep duration (right)"}
p1 <- ggplot(df_model, aes(term_gpa)) +
  geom_histogram(aes(fill = after_stat(x) > 3.0), bins = 30, color = "white") +
  scale_fill_manual(values = c("FALSE" = "#AAAAAA", "TRUE" = "#C8102E"),
                    labels = c("< 3.0", "≥ 3.0"), name = "GPA") +
  theme_minimal(base_size = 12) +
  labs(title = "Term GPA Distribution", x = "Term GPA", y = "Count")

p2 <- ggplot(df_model, aes(TotalSleepTime)) +
  geom_density(fill = "#2166AC", alpha = 0.45, color = "white") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "#C8102E",
             linewidth = 0.9) +
  annotate("text", x = 7.2, y = 0.35, label = "7h reference",
           hjust = 0, color = "#C8102E", size = 3.5) +
  theme_minimal(base_size = 12) +
  labs(title = "Sleep Duration Distribution",
       x = "Total Sleep (hours)", y = "Density")

patchwork::wrap_plots(p1, p2)
```

::: fill-box
GPA is left-skewed with most students above 3.0. Sleep clusters around 6–8
hours, with the majority below the recommended 7-hour threshold.
:::

## Sleep vs GPA Scatterplot

```{r sleep-gpa-scatter, fig.cap="Figure 2: Scatterplot of sleep duration vs term GPA with regression line"}
ggplot(df_model, aes(TotalSleepTime, term_gpa, color = high_gpa)) +
  geom_point(alpha = 0.45, size = 1.8) +
  geom_smooth(aes(group = 1), method = "lm", color = "#C8102E",
              linewidth = 1.1, se = TRUE) +
  scale_color_manual(values = c("0" = "#AAAAAA", "1" = "#2166AC"),
                     labels = c("Low GPA", "High GPA"), name = NULL) +
  theme_minimal(base_size = 12) +
  labs(title = "Sleep Duration vs Term GPA",
       subtitle = "Color indicates high (≥3.0) vs low GPA; OLS regression line shown",
       x = "Total Sleep Time (hours)", y = "Term GPA")
```

## Sleep Category & High GPA

```{r sleep-cat-bar, fig.height=4.5, fig.cap="Figure 3: Proportion of high GPA students by sleep category"}
p_prop <- df_model %>%
  count(sleep_cat, high_gpa) %>%
  group_by(sleep_cat) %>%
  mutate(prop = n / sum(n)) %>%
  filter(high_gpa == 1) %>%
  ggplot(aes(sleep_cat, prop, fill = sleep_cat)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = percent(prop, accuracy = 1)), vjust = -0.4, size = 4) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Proportion with High GPA by Sleep Category",
       x = "Sleep Category", y = "Proportion (High GPA)")

p_count <- df_model %>%
  count(sleep_cat, high_gpa) %>%
  ggplot(aes(sleep_cat, n, fill = high_gpa)) +
  geom_col(position = "fill", width = 0.55) +
  scale_fill_manual(values = c("0" = "#AAAAAA", "1" = "#C8102E"),
                    labels = c("Low GPA", "High GPA"), name = NULL) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal(base_size = 12) +
  labs(title = "GPA Class by Sleep Category",
       x = "Sleep Category", y = "Proportion")

p_prop | p_count
```

## GPA by Study Group

```{r gpa-by-study, fig.cap="Figure 4: Term GPA distribution by university study group"}
group_stats <- df_model %>%
  group_by(study) %>%
  summarise(mean_gpa = mean(term_gpa, na.rm = TRUE),
            se_gpa   = sd(term_gpa, na.rm = TRUE) / sqrt(n()),
            n        = n())

ggplot(df_model, aes(study, term_gpa, fill = study)) +
  geom_boxplot(alpha = 0.5, outlier.shape = 21, outlier.size = 1.5) +
  geom_point(data = group_stats, aes(study, mean_gpa),
             shape = 18, size = 4, color = "#C8102E") +
  geom_errorbar(data = group_stats,
                aes(x = study, y = mean_gpa,
                    ymin = mean_gpa - 1.96 * se_gpa,
                    ymax = mean_gpa + 1.96 * se_gpa),
                width = 0.15, color = "#C8102E", linewidth = 0.8,
                inherit.aes = FALSE) +
  scale_fill_viridis_d(option = "C", begin = 0.1, end = 0.9) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Term GPA by University Study Group",
       subtitle = "Diamond = group mean; error bars = 95% CI",
       x = "Study Group", y = "Term GPA")
```

## Correlation Structure

```{r correlation-heatmap, fig.cap="Figure 5: Correlation matrix of numeric predictors"}
num_vars <- df_model %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ) %>%
  rename(
    "Term GPA"    = term_gpa,
    "Sleep (hrs)" = TotalSleepTime,
    "Cum GPA"     = cum_gpa,
    "Course Load" = Zterm_units_ZofZ
  )

cor_mat <- cor(num_vars, use = "complete.obs")

corrplot(cor_mat, method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black",
         tl.srt = 45,
         col = colorRampPalette(c("#2166AC", "white", "#C8102E"))(200),
         na.label = " ",
         title    = "Correlation Matrix",
         mar = c(0, 0, 2, 0))
```

## Pairwise Relationships

```{r ggpairs, fig.height=6, fig.cap="Figure 6: Pairwise variable relationships"}
ggpairs(
  num_vars,
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8,
               color = "#2166AC")),
  diag  = list(continuous = wrap("densityDiag", fill = "#C8102E", alpha = 0.4)),
  upper = list(continuous = wrap("cor", size = 4))
) +
  theme_minimal(base_size = 10) +
  labs(title = "Pairwise Variable Relationships")
```

::: fill-box
**EDA Takeaways**

- Cumulative GPA shows the strongest correlation with term GPA
- Sleep duration has a modest but positive correlation with GPA
- Course load shows a weak relationship
- Students sleeping >8h have the highest proportion of high GPA
:::

---

# Multiple Linear Regression

**Research Question:** Does sleep duration significantly affect term GPA after
controlling for cumulative GPA and course load?

**Hypothesis:** H₀: β_sleep = 0 vs Hₐ: β_sleep ≠ 0

## Model Fitting

```{r mlr-model}
mlr <- lm(term_gpa ~ TotalSleepTime + cum_gpa + Zterm_units_ZofZ,
          data = df_model)

# Regression equation coefficients
tidy(mlr, conf.int = TRUE) %>%
  mutate(sig = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01  ~ "**",
    p.value < 0.05  ~ "*",
    TRUE            ~ ""
  )) %>%
  kable(digits = 4, caption = "Table 3: MLR Coefficients")
```

```{r mlr-fit}
# Model fit statistics
glance(mlr) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, nobs) %>%
  kable(digits = 4, caption = "Table 4: Model Fit Statistics")
```

::: result-box
**Regression Equation:**
term_gpa = 0.39 + 0.10 × Sleep(hrs) + 0.67 × cum_gpa + 0.03 × CourseLoad

**Interpretation (holding all else constant):**

- Each additional hour of sleep is associated with a **0.10-point increase** in GPA
- Each 1-point increase in cumulative GPA is associated with a **0.67-point increase** in term GPA
- Course load is **not statistically significant** (p > 0.05)
- Model explains approximately **40%** of variance in GPA (R² ≈ 0.41)

The positive coefficient for sleep suggests a linear relationship, consistent with patterns observed in the exploratory analysis.
:::

## Coefficient Plot

```{r coef-plot, fig.height=4, fig.cap="Figure 7: MLR coefficient estimates with 95% confidence intervals"}
tidy(mlr, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "TotalSleepTime"   ~ "Sleep (hrs)",
      term == "cum_gpa"          ~ "Cumulative GPA",
      term == "Zterm_units_ZofZ" ~ "Course Load (Z)",
      TRUE ~ term),
    significant = p.value < 0.05
  ) %>%
  ggplot(aes(estimate, fct_reorder(term, estimate), color = significant)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#888888") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("FALSE" = "#AAAAAA", "TRUE" = "#C8102E"),
                     labels = c("Not significant", "p < 0.05"), name = NULL) +
  theme_minimal(base_size = 12) +
  labs(title = "MLR Coefficient Plot (95% CI)",
       subtitle = "Red = statistically significant at α = 0.05",
       x = "Coefficient Estimate", y = NULL)
```

## Multicollinearity (VIF)

```{r vif}
vif(mlr) %>%
  enframe(name = "Variable", value = "VIF") %>%
  mutate(Concern = if_else(VIF > 5, "High", "OK")) %>%
  kable(digits = 3, caption = "Table 5: Variance Inflation Factors")
# All VIF < 2: no multicollinearity concern
```

## Diagnostic Plots

```{r mlr-diagnostics, fig.height=6, fig.cap="Figure 8: Regression diagnostic plots (LINE assumption checks)"}
diag_df <- data.frame(
  fitted    = fitted(mlr),
  residuals = residuals(mlr),
  std_resid = rstandard(mlr),
  leverage  = hatvalues(mlr)
)

p_rv <- ggplot(diag_df, aes(fitted, residuals)) +
  geom_point(alpha = 0.4, color = "#2166AC") +
  geom_hline(yintercept = 0, color = "#C8102E", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "#C8102E", linewidth = 0.8, method = "loess") +
  theme_minimal(base_size = 11) +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals")

p_qq <- ggplot(diag_df, aes(sample = std_resid)) +
  stat_qq(alpha = 0.5, color = "#2166AC") +
  stat_qq_line(color = "#C8102E", linewidth = 0.9) +
  theme_minimal(base_size = 11) +
  labs(title = "Normal Q-Q Plot",
       x = "Theoretical Quantiles", y = "Std. Residuals")

p_sl <- ggplot(diag_df, aes(fitted, sqrt(abs(std_resid)))) +
  geom_point(alpha = 0.4, color = "#2166AC") +
  geom_smooth(se = FALSE, color = "#C8102E", linewidth = 0.8, method = "loess") +
  theme_minimal(base_size = 11) +
  labs(title = "Scale-Location", x = "Fitted Values", y = "√|Std. Residuals|")

p_lv <- ggplot(diag_df, aes(leverage, std_resid)) +
  geom_point(alpha = 0.4, color = "#2166AC") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "#C8102E") +
  theme_minimal(base_size = 11) +
  labs(title = "Leverage vs Std. Residuals",
       x = "Leverage", y = "Std. Residuals")

(p_rv | p_qq) / (p_sl | p_lv)
```

::: fill-box
**LINE Assumption Checks:**

- **Linearity**: Residuals vs Fitted shows no strong pattern - assumption holds
- **Independence**: Observations are independent students from different cohorts
- **Normality**: Q-Q plot shows approximate normality with mild tail deviations
- **Equal Variance**: Scale-Location plot is roughly flat - homoscedasticity satisfied
- No extreme influential observations in the leverage plot
:::

::: result-box
**Answer to RQ1:** Yes, sleep duration significantly affects term GPA (β = 0.10,
p < 0.001) after controlling for cumulative GPA and course load. Each additional
hour of sleep is associated with approximately a 0.10-point increase in GPA,
holding other variables constant. However, cumulative GPA remains the dominant
predictor (β = 0.67).
:::

---

# One-Way ANOVA

**Research Question:** Does mean GPA differ across study groups (universities)?

**Hypothesis:** H₀: μ₁ = μ₂ = μ₃ = μ₄ vs Hₐ: Not all group means are equal

## ANOVA F-Test & Effect Size

```{r anova-model}
anova_model <- aov(term_gpa ~ study, data = df_model)
summary(anova_model)

# Eta-squared effect size
anova_tbl <- summary(anova_model)[[1]]
eta_sq    <- anova_tbl["study", "Sum Sq"] / sum(anova_tbl[, "Sum Sq"])
cat(sprintf("Eta-squared: %.4f (%.1f%% of variance explained by study group)\n",
            eta_sq, 100 * eta_sq))
```

```{r anova-table}
broom::tidy(anova_model) %>%
  kable(digits = 4, caption = "Table 6: One-Way ANOVA Results")
```

## Assumption Checks

```{r anova-assumptions}
# Normality of residuals
shapiro.test(residuals(anova_model))
# Although residuals deviate from normality (p < 0.001), 
# ANOVA is robust to such violations given the large sample size (n = 487)

# Equality of variances
leveneTest(term_gpa ~ study, data = df_model)
# p > 0.05: equal variance assumption satisfied
```

## Post-Hoc: Tukey HSD

```{r tukey-table}
tukey_res <- TukeyHSD(anova_model)

as.data.frame(tukey_res$study) %>%
  rownames_to_column("Comparison") %>%
  rename(Estimate = diff, Lower = lwr, Upper = upr, `p adj` = `p adj`) %>%
  mutate(Significant = if_else(`p adj` < 0.05, "Yes", "No")) %>%
  kable(digits = 4, caption = "Table 7: Tukey HSD Post-Hoc Comparisons")
```

```{r tukey-ci-plot, fig.height=5, fig.cap="Figure 9: Tukey HSD 95% confidence intervals for pairwise differences"}
# Tukey HSD confidence interval plot (ggplot version - avoids base R 'main' bug)
tukey_df <- as.data.frame(tukey_res$study) %>%
  rownames_to_column("Comparison") %>%
  rename(Estimate = diff, Lower = lwr, Upper = upr, p_adj = `p adj`) %>%
  mutate(Significant = if_else(p_adj < 0.05, "Yes", "No"))

ggplot(tukey_df, aes(x = Estimate, y = fct_reorder(Comparison, Estimate),
                     color = Significant)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#888888") +
  geom_errorbarh(aes(xmin = Lower, xmax = Upper), height = 0.25,
                 linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("No" = "#AAAAAA", "Yes" = "#C8102E"),
                     name = "p < 0.05") +
  theme_minimal(base_size = 12) +
  labs(title = "Tukey HSD 95% Family-Wise Confidence Intervals",
       subtitle = "Intervals crossing zero = not significant",
       x = "Difference in Mean GPA", y = NULL)
```

## ANOVA Visualization

```{r anova-viz, fig.height=5, fig.cap="Figure 10: Mean term GPA by study group with 95% CI"}
group_means <- df_model %>%
  group_by(study) %>%
  summarise(
    mean_gpa = mean(term_gpa, na.rm = TRUE),
    se       = sd(term_gpa, na.rm = TRUE) / sqrt(n()),
    n        = n()
  )

ggplot(group_means, aes(fct_reorder(study, mean_gpa), mean_gpa, fill = study)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_gpa - 1.96 * se, ymax = mean_gpa + 1.96 * se),
                width = 0.2, linewidth = 0.8) +
  geom_text(aes(label = paste0(round(mean_gpa, 2), "\n(n=", n, ")")),
            vjust = -0.6, size = 3.5) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8) +
  scale_y_continuous(limits = c(0, 4)) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Mean Term GPA by Study Group",
       subtitle = "Error bars = 95% CI",
       x = NULL, y = "Mean Term GPA")
```

::: result-box
**Answer to RQ2:** Yes, mean GPA differs significantly across study groups,
F(3, 483) = 5.02, p = 0.002. Effect size η² ≈ 0.03 indicates study group
explains approximately 3% of GPA variability - a small but statistically
meaningful effect. Tukey HSD reveals three significant pairwise differences:
Study 5 vs 1 (p = 0.041), Study 3 vs 2 (p = 0.048), and Study 5 vs 2
(p = 0.006). Groups 1 and 2 tend to have higher mean GPA than groups 3 and 5.
:::

---

# Logistic Regression

**Research Question:** How do sleep duration, cumulative GPA, and course load
predict whether a student achieves a high GPA (≥ 3.0)?

**Model:** log(P(high_gpa=1) / (1 - P(high_gpa=1))) = β₀ + β₁(Sleep) + β₂(cum_gpa) + β₃(CourseLoad)

## Model Fitting

```{r logistic-model}
logit <- glm(high_gpa ~ TotalSleepTime + cum_gpa + Zterm_units_ZofZ,
             data = df_model, family = binomial)
summary(logit)
```

## Odds Ratios with Confidence Intervals

```{r odds-ratios}
exp(cbind(OR = coef(logit), confint.default(logit))) %>%
  as.data.frame() %>%
  rownames_to_column("Term") %>%
  filter(Term != "(Intercept)") %>%
  rename(Lower_95 = `2.5 %`, Upper_95 = `97.5 %`) %>%
  kable(digits = 4,
        caption = "Table 8: Odds Ratios (exponentiated coefficients)")
```

::: result-box
**Odds Ratio Interpretation (holding all else constant):**

- **Sleep:** OR = 1.64 -  Each additional hour of sleep is associated with a 64% increase in the odds of achieving a high GPA, holding other variables constant.
- **Cum GPA:** OR = 15.65 - Each 1-point increase in cumulative GPA multiplies the odds of achieving a high GPA by approximately 15.7.
- **Course Load:** OR ≈ 0.97 - not significant (CI includes 1.0)
:::

## Predicted Probability vs Sleep

```{r logistic-viz, fig.height=4.5, fig.cap="Figure 11: Predicted probability of high GPA vs sleep duration"}
df_model_logit <- df_model %>%
  mutate(pred_prob = predict(logit, type = "response"))

ggplot(df_model_logit, aes(TotalSleepTime, pred_prob, color = high_gpa)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(aes(group = 1), method = "loess", color = "#C8102E",
              linewidth = 1, se = TRUE) +
  scale_color_manual(values = c("0" = "#AAAAAA", "1" = "#2166AC"),
                     labels = c("Low GPA", "High GPA"), name = "Actual") +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal(base_size = 12) +
  labs(title = "Predicted Probability of High GPA vs Sleep Duration",
       x = "Total Sleep Time (hours)", y = "P(High GPA)")
```

## ROC Curve & AUC

```{r roc-curve, fig.height=5, fig.cap="Figure 12: ROC curve for logistic regression model"}

roc_obj <- roc(
  response  = df_model_logit$high_gpa,
  predictor = df_model_logit$pred_prob,
  levels    = c("0", "1"),
  quiet     = TRUE
)

auc_val <- round(as.numeric(auc(roc_obj)), 4)

roc_df <- data.frame(
  specificity = roc_obj$specificities,
  sensitivity = roc_obj$sensitivities
)

ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
  geom_line(color = "#2166AC", linewidth = 1.2) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "#999999") +
  
  annotate("text",
           x = 0.75, y = 0.15,
           label = paste0("AUC = ", auc_val),
           size = 4.5, color = "#C8102E") +
  
  coord_equal() +
  theme_minimal(base_size = 12) +
  labs(
    title = "ROC Curve - Logistic Regression",
    subtitle = "Predicting High GPA (≥ 3.0)",
    x = "False Positive Rate",
    y = "True Positive Rate"
  )

```

## Confusion Matrix

```{r confusion-matrix}
pred_class <- factor(if_else(df_model_logit$pred_prob >= 0.5, "1", "0"),
                     levels = levels(df_model_logit$high_gpa))
cm <- confusionMatrix(pred_class, df_model_logit$high_gpa, positive = "1")
cm
```

## Threshold Analysis

```{r threshold-table}
thresholds <- seq(0.3, 0.7, by = 0.05)

threshold_tbl <- map_df(thresholds, function(th) {
  pred <- factor(if_else(df_model_logit$pred_prob >= th, "1", "0"),
                 levels = c("0", "1"))
  cm_t <- confusionMatrix(pred, df_model_logit$high_gpa, positive = "1")
  tibble(
    Threshold   = th,
    Accuracy    = cm_t$overall["Accuracy"],
    Sensitivity = cm_t$byClass["Sensitivity"],
    Specificity = cm_t$byClass["Specificity"],
    Precision   = cm_t$byClass["Pos Pred Value"]
  )
})

threshold_tbl %>%
  kable(digits = 3, caption = "Table 9: Logistic Regression at Different Thresholds")
```

```{r threshold-tradeoff-plot, fig.height=4.5, fig.cap="Figure 13: Sensitivity/specificity/accuracy tradeoff across thresholds"}
# NEW: Threshold tradeoff visualization
threshold_tbl %>%
  pivot_longer(cols = c(Accuracy, Sensitivity, Specificity, Precision),
               names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(Threshold, Value, color = Metric)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "#888888") +
  annotate("text", x = 0.52, y = 0.35, label = "Default (0.5)",
           hjust = 0, size = 3, color = "#888888") +
  scale_color_manual(values = c(
    "Accuracy"    = "#2166AC",
    "Sensitivity" = "#C8102E",
    "Specificity" = "#22863a",
    "Precision"   = "#6f42c1"
  )) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  theme_minimal(base_size = 12) +
  labs(title = "Classification Metrics Across Decision Thresholds",
       subtitle = "Threshold of 0.5 provides reasonable sensitivity-specificity balance",
       x = "Decision Threshold", y = "Metric Value")
```

::: result-box
**Answer to RQ3:** Sleep duration positively influences the probability of
achieving a high GPA (OR = 1.64, p < 0.05), but cumulative GPA remains the
dominant predictor (OR = 15.65). The model achieves AUC = `r auc_val`,
indicating good discrimination. At threshold = 0.5: accuracy ≈ 84%, sensitivity
≈ 96%, specificity ≈ 31%. The low specificity reflects class imbalance (~82%
of students have high GPA). Threshold 0.5 is retained as a reasonable
compromise. The ROC curve demonstrates strong discrimination between high and low GPA students despite class imbalance in the dataset.
:::

---

# Machine Learning Pipeline (caret)

Following the industry-standard caret framework: **split → preprocess →
cross-validate → compare → evaluate**.

## Train/Test Split

```{r ml-split, fig.height=3.5}
set.seed(6105)

df_ml <- df_model %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ) %>%
  drop_na()

train_idx  <- createDataPartition(df_ml$term_gpa, p = 0.80, list = FALSE)
train_data <- df_ml[ train_idx, ]
test_data  <- df_ml[-train_idx, ]

cat(sprintf("Training observations: %d | Test observations: %d\n",
            nrow(train_data), nrow(test_data)))

bind_rows(
  train_data %>% mutate(split = "Train"),
  test_data  %>% mutate(split = "Test")
) %>%
  ggplot(aes(term_gpa, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Train" = "#2166AC", "Test" = "#C8102E")) +
  theme_minimal(base_size = 12) +
  labs(title = "GPA Distribution: Train vs Test Split",
       x = "Term GPA", y = "Density", fill = "Split")
```

## Cross-Validation & Model Zoo

```{r cv-control}
ctrl <- trainControl(
  method          = "cv",
  number          = 10,
  savePredictions = "final",
  verboseIter     = FALSE
)
```

```{r model-zoo, cache=TRUE}
set.seed(6105)

# Linear Regression
m_lm <- train(term_gpa ~ ., data = train_data, method = "lm",
              preProcess = c("center", "scale"), trControl = ctrl)

# K-Nearest Neighbours
m_knn <- train(term_gpa ~ ., data = train_data, method = "knn",
               preProcess = c("center", "scale"), trControl = ctrl,
               tuneGrid = data.frame(k = c(5, 7, 9, 11, 15, 21)))

# Decision Tree
m_tree <- train(term_gpa ~ ., data = train_data, method = "rpart",
                preProcess = c("center", "scale"), trControl = ctrl,
                tuneLength = 10)

# Random Forest
m_rf <- train(term_gpa ~ ., data = train_data, method = "rf",
              preProcess = c("center", "scale"), trControl = ctrl,
              tuneGrid = data.frame(mtry = c(1, 2, 3)),
              ntree = 200)

# Ridge Regression
m_ridge <- train(term_gpa ~ ., data = train_data, method = "glmnet",
                 preProcess = c("center", "scale"), trControl = ctrl,
                 tuneGrid = expand.grid(alpha = 0,
                                        lambda = 10^seq(-3, 1, length = 30)))

# Lasso Regression
m_lasso <- train(term_gpa ~ ., data = train_data, method = "glmnet",
                 preProcess = c("center", "scale"), trControl = ctrl,
                 tuneGrid = expand.grid(alpha = 1,
                                        lambda = 10^seq(-3, 1, length = 30)))
```

## Cross-Validation Comparison

```{r model-comparison}
resamps <- resamples(list(
  `Linear`        = m_lm,
  `KNN`           = m_knn,
  `Decision Tree` = m_tree,
  `Random Forest` = m_rf,
  `Ridge`         = m_ridge,
  `Lasso`         = m_lasso
))

summary(resamps)$statistics$RMSE %>%
  as.data.frame() %>%
  rownames_to_column("Model") %>%
  arrange(Mean) %>%
  kable(digits = 4, caption = "Table 10: 10-Fold CV RMSE Across Models")
```

```{r dotplot-cv, fig.height=5, fig.cap="Figure 14: 10-fold CV RMSE comparison across models"}
dotplot(resamps, metric = "RMSE",
        main = "10-Fold CV RMSE - Model Comparison")
```

```{r bwplot-cv, fig.height=5, fig.cap="Figure 15: 10-fold CV R-squared comparison across models"}
bwplot(resamps, metric = "Rsquared",
       main = "10-Fold CV R² - Model Comparison")
```

## Test Set Evaluation

```{r test-evaluation}
eval_model <- function(model, name) {
  preds <- predict(model, newdata = test_data)
  pr    <- postResample(pred = preds, obs = test_data$term_gpa)
  tibble(Model = name, RMSE = pr["RMSE"], MAE = pr["MAE"],
         Rsquared = pr["Rsquared"])
}

test_results <- bind_rows(
  eval_model(m_lm,    "Linear Regression"),
  eval_model(m_knn,   "KNN"),
  eval_model(m_tree,  "Decision Tree"),
  eval_model(m_rf,    "Random Forest"),
  eval_model(m_ridge, "Ridge Regression"),
  eval_model(m_lasso, "Lasso Regression")
) %>%
  arrange(RMSE)

test_results %>%
  kable(digits = 4, caption = "Table 11: Test Set Performance")
```

## Actual vs Predicted (Best Model)

```{r actual-vs-predicted, fig.height=5, fig.cap="Figure 16: Actual vs predicted GPA for best model"}
best_model_name <- test_results$Model[1]
best_model_obj  <- list(
  "Linear Regression" = m_lm,
  "KNN"               = m_knn,
  "Decision Tree"     = m_tree,
  "Random Forest"     = m_rf,
  "Ridge Regression"  = m_ridge,
  "Lasso Regression"  = m_lasso
)[[best_model_name]]

pred_all <- tibble(
  Actual    = test_data$term_gpa,
  Predicted = predict(best_model_obj, newdata = test_data),
  Residual  = Actual - Predicted
)

p_avp <- ggplot(pred_all, aes(Actual, Predicted)) +
  geom_point(aes(color = abs(Residual)), alpha = 0.65, size = 2) +
  geom_abline(linetype = "dashed", color = "#C8102E", linewidth = 1) +
  scale_color_viridis_c(option = "C", name = "|Residual|") +
  theme_minimal(base_size = 12) +
  labs(title = paste("Actual vs Predicted -", best_model_name),
       subtitle = "Dashed line = perfect prediction",
       x = "Actual Term GPA", y = "Predicted Term GPA")

p_res <- ggplot(pred_all, aes(Predicted, Residual)) +
  geom_point(aes(color = abs(Residual)), alpha = 0.65, size = 2) +
  geom_hline(yintercept = 0, color = "#C8102E", linetype = "dashed") +
  scale_color_viridis_c(option = "C", name = "|Residual|") +
  theme_minimal(base_size = 12) +
  labs(title = "Residuals vs Predicted",
       x = "Predicted Term GPA", y = "Residual")

p_avp | p_res
```

## Variable Importance

```{r variable-importance, fig.height=4, fig.cap="Figure 17: Variable importance from Random Forest model"}
vi_rf <- varImp(m_rf)$importance %>%
  rownames_to_column("Variable") %>%
  arrange(desc(Overall)) %>%
  mutate(Variable = case_when(
    Variable == "cum_gpa"          ~ "Cumulative GPA",
    Variable == "TotalSleepTime"   ~ "Sleep Duration (hrs)",
    Variable == "Zterm_units_ZofZ" ~ "Course Load (Z)",
    TRUE ~ Variable
  ))

ggplot(vi_rf, aes(Overall, fct_reorder(Variable, Overall),
                  fill = Variable == "Sleep Duration (hrs)")) +
  geom_col(width = 0.55) +
  scale_fill_manual(values = c("FALSE" = "#AAAAAA", "TRUE" = "#C8102E"),
                    guide = "none") +
  theme_minimal(base_size = 12) +
  labs(title = "Variable Importance - Random Forest",
       subtitle = "Sleep highlighted in red",
       x = "Importance Score", y = NULL)
```

::: result-box
**ML Pipeline Findings:**

- Cumulative GPA dominates as the most important predictor across all models
- Linear and regularized models perform competitively; complex models (KNN,
  Decision Tree) do not provide major gains
- This confirms that the sleep-GPA relationship is largely linear
- Best model: `r best_model_name` (Test RMSE = `r round(test_results$RMSE[1], 4)`)
:::

---

# Nonparametric Validation

**Research Question:** Are the GPA group differences and sleep-GPA association
robust when distributional assumptions are relaxed?

**Justification:** The Shapiro-Wilk test (p < 0.001) indicated non-normality of
ANOVA residuals. Nonparametric methods do not assume normal distributions,
providing a robustness check.

```{r nonparametric}
# Kruskal-Wallis: nonparametric alternative to ANOVA
kw <- kruskal.test(term_gpa ~ study, data = df_model)

# Spearman correlation: nonparametric alternative to Pearson
sp <- cor.test(df_model$TotalSleepTime, df_model$term_gpa,
               method = "spearman", use = "complete.obs")

# Comparison table
tibble(
  Test = c("Kruskal-Wallis (study → GPA)", "Spearman (sleep ~ GPA)"),
  Statistic = c(round(kw$statistic, 3), round(sp$estimate, 4)),
  `p-value`  = c(format.pval(kw$p.value, digits = 3),
                 format.pval(sp$p.value, digits = 3)),
  `Parametric Equivalent` = c(
    paste0("ANOVA: F = 5.02, p = 0.002"),
    paste0("Pearson: r = ", round(cor(df_model$TotalSleepTime,
           df_model$term_gpa), 4))
  ),
  Conclusion = c(
    if_else(kw$p.value < 0.05,
            "Confirms ANOVA: groups differ",
            "Contradicts ANOVA"),
    if_else(sp$p.value < 0.05,
            "Confirms positive sleep-GPA association",
            "No significant association")
  )
) %>%
  kable(caption = "Table 12: Nonparametric Validation - Comparison with Parametric Results")
```

::: result-box
**Answer to RQ4:** Yes. Kruskal-Wallis (χ² = `r round(kw$statistic, 2)`,
p = `r format.pval(kw$p.value, digits = 3)`) confirms that GPA differs
significantly across study groups, consistent with the parametric ANOVA.
Spearman correlation (ρ = `r round(sp$estimate, 3)`,
p = `r format.pval(sp$p.value, digits = 3)`) confirms a significant positive
monotonic association between sleep and GPA. Both nonparametric results validate
the parametric findings, demonstrating that conclusions are robust to
distributional assumptions.
:::

---

# Summary of Findings

```{r findings-table, echo=FALSE}
tibble(
  Method = c("Multiple Linear Regression", "One-Way ANOVA",
             "Logistic Regression", "ML Pipeline (caret)",
             "Nonparametric Tests"),
  `Key Finding` = c(
    paste0("R² = ", round(glance(mlr)$r.squared, 3),
           "; cum_gpa strongest (β=0.67); sleep significant (β=0.10, p<0.001)"),
    paste0("F(3,483) = 5.02, p = 0.002; η² = ", round(eta_sq, 3),
           "; 3 significant pairwise differences"),
    paste0("AUC = ", auc_val,
           "; each hour of sleep → 64% higher odds of high GPA"),
    paste0("Best: ", best_model_name, " (RMSE = ",
           round(test_results$RMSE[1], 4),
           "); linear models outperform complex models"),
    paste0("Kruskal-Wallis p = ", format.pval(kw$p.value, digits = 3),
           "; Spearman ρ = ", round(sp$estimate, 3), " - confirms all findings")
  )
) %>%
  kable(caption = "Table 13: Summary of Statistical Findings Across All Methods")
```

---

# Conclusion

- **Cumulative GPA** is by far the strongest predictor of term GPA across every
  method tested (β = 0.67, OR = 15.65, highest variable importance)
- **Sleep duration** shows a consistent, positive, and statistically significant
  but modest association with academic performance (β = 0.10, OR = 1.64,
  Spearman ρ > 0)
- Students sleeping **more than 8 hours** have the highest proportion of high
  GPAs, while those sleeping **less than 6 hours** have the lowest
- **GPA varies significantly by study group** (p = 0.002, η² = 0.03),
  suggesting institutional or cohort-level factors play a role
- **All five analytical approaches converge** on the same general conclusion,
  strengthening confidence in these findings
- Simpler linear models match or outperform complex ML models, confirming
  that the underlying relationships are primarily linear
- This suggests that the relationship between predictors and GPA is predominantly linear, with limited benefit from nonlinear modeling approaches
- These results are consistent across parametric, nonparametric, and machine learning approaches, strengthening the robustness of the findings

::: concept-box
**Practical Implication:** Encouraging adequate sleep (≥ 7 hours) may
contribute to better academic outcomes, but prior academic performance is the
strongest single driver. Sleep improvement should be considered one component
of a broader academic support strategy, rather than a standalone intervention.
:::

---

# Limitations

Several limitations should be considered when interpreting these findings:

- **Observational design:** The data are observational, so the analyses
  identify associations rather than causal effects. Although sleep duration
  is positively related to GPA, these results do not prove that increasing
  sleep will directly improve academic performance.

- **Measurement error:** Sleep was measured using wearable Fitbit devices,
  which may introduce error through imperfect accuracy, missed wear time,
  or incomplete nightly capture. Such error could attenuate the estimated
  sleep–GPA relationship.

- **Missing data:** The course load variable (Zterm_units_ZofZ) has ~23.2%
  missing values. Complete-case analysis reduces the sample from 634 to 487
  and may introduce bias if missingness is not random.

- **Limited generalizability:** The dataset includes only first-year students
  from three U.S. universities. Findings may not generalize to upper-year
  students, other institutions, or non-U.S. populations.

- **Unmeasured confounders:** Important factors such as stress, motivation,
  study habits, mental health, and social support are not captured in the
  dataset. These omitted variables may partly explain both sleep patterns
  and academic performance, potentially confounding the observed
  association.

---

# References

1. CMU S&DS Data Repository - Nightly Sleep Time and GPA in First-Year Students
   (2023 dataset, publicly available).
   https://cmustatistics.github.io/data-repository/

2. Creswell, J. D., et al. (2023). Nightly sleep duration predicts grade point
   average in the first year of college. *Proceedings of the National Academy
   of Sciences*.

3. Kuhn, M. (2008). Building predictive models in R using the caret package.
   *Journal of Statistical Software*.

4. Robin, X., et al. (2011). pROC: An open-source package for R and S+ to
   analyze and compare ROC curves. *BMC Bioinformatics*.

5. Fox, J., & Weisberg, S. (2019). *An R Companion to Applied Regression*.
   Sage Publications.

6. Wickham, H., et al. (2019). Welcome to the tidyverse. *Journal of Open
   Source Software*.

---

```{r session-info, echo=FALSE}
sessionInfo()
```
