---
title: "INFO 6105 Final Project"
subtitle: "Impact of Sleep Duration on Academic Performance (GPA)"
author: "Rushita Vachhani"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5.5)

library(tidyverse)
library(broom)
library(car)
library(pROC)
library(caret)
library(here)
library(GGally)
library(corrplot)
library(patchwork)
library(viridis)
library(knitr)
library(scales)
library(purrr)
library(randomForest)
```

```{=html}
<style>
body { font-family: 'Palatino', 'Georgia', serif; font-size: 15px; }
h1, h2 { color: #C8102E; border-bottom: 2px solid #C8102E; padding-bottom: 4px; }
h3 { color: #333333; }
.fill-box {
  border: 2px solid #C8102E; padding: 12px; margin: 10px 0;
  background-color: #fdf5f5; border-radius: 5px;
}
.concept-box {
  border: 2px solid #2166AC; padding: 12px; margin: 10px 0;
  background-color: #f0f6fc; border-radius: 5px;
}
.result-box {
  border: 2px solid #22863a; padding: 12px; margin: 10px 0;
  background-color: #f0fff4; border-radius: 5px;
}
</style>
```

---

# Introduction

This project investigates whether **sleep duration impacts academic performance (GPA)** using real student wearable + academic data from Carnegie Mellon University.

We apply multiple statistical methods — Multiple Linear Regression, ANOVA, Logistic Regression, and a full Machine Learning pipeline — to rigorously evaluate this relationship while controlling for cumulative GPA and course load.

::: concept-box
**Research Question:** Does total sleep time predict end-of-term GPA, and how does it compare to other academic predictors?
:::

---

# Data Preparation

```{r load-data}
df_raw <- read.csv(here("data/raw/cmu-sleep.csv"))

df <- df_raw %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ, study) %>%
  mutate(
    TotalSleepTime = TotalSleepTime / 60,          # minutes -> hours
    high_gpa       = factor(if_else(term_gpa >= 3.0, 1, 0), levels = c(0, 1)),
    study          = factor(study),
    sleep_cat      = factor(
      case_when(
        TotalSleepTime < 6  ~ "Low (<6h)",
        TotalSleepTime <= 8 ~ "Medium (6–8h)",
        TRUE                ~ "High (>8h)"
      ),
      levels = c("Low (<6h)", "Medium (6–8h)", "High (>8h)")
    )
  )

# Modelling subset: complete cases
df_model <- df %>%
  drop_na(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ, study, high_gpa)
```

## Dataset Overview

```{r dataset-summary}
tibble(
  Variable       = c("term_gpa", "TotalSleepTime (hrs)", "cum_gpa", "Zterm_units_ZofZ", "study (groups)"),
  N_complete     = c(sum(!is.na(df$term_gpa)), sum(!is.na(df$TotalSleepTime)),
                     sum(!is.na(df$cum_gpa)),  sum(!is.na(df$Zterm_units_ZofZ)),
                     sum(!is.na(df$study))),
  Missing        = c(sum(is.na(df$term_gpa)), sum(is.na(df$TotalSleepTime)),
                     sum(is.na(df$cum_gpa)),  sum(is.na(df$Zterm_units_ZofZ)),
                     sum(is.na(df$study)))
) %>%
  kable(caption = "Missingness Summary")
```

```{r summary-stats}
df_model %>%
  summarise(
    across(c(term_gpa, TotalSleepTime, cum_gpa),
           list(Mean = ~round(mean(.),2), SD = ~round(sd(.),2),
                Min  = ~round(min(.),2),  Max = ~round(max(.),2)),
           .names = "{.col}_{.fn}")
  ) %>%
  pivot_longer(everything(), names_to = c("Variable", "Stat"), names_sep = "_(?=[^_]+$)") %>%
  pivot_wider(names_from = Stat, values_from = value) %>%
  kable(caption = "Descriptive Statistics")
```

::: fill-box
**Key Preparation Steps**

- Sleep time converted from minutes to hours
- Binary outcome `high_gpa` created (1 = term GPA ≥ 3.0)
- Sleep categories: Low < 6 h, Medium 6–8 h, High > 8 h
- Complete-case dataset used for all models: **`r nrow(df_model)` observations**
:::

---

# Exploratory Data Analysis

## Outcome & Predictor Distributions

```{r eda-distributions, fig.height=5}
p1 <- ggplot(df_model, aes(term_gpa)) +
  geom_histogram(aes(fill = after_stat(x) > 3.0), bins = 30, color = "white") +
  scale_fill_manual(values = c("FALSE" = "#AAAAAA", "TRUE" = "#C8102E"),
                    labels = c("< 3.0", "≥ 3.0"), name = "GPA") +
  theme_minimal(base_size = 12) +
  labs(title = "Term GPA Distribution", x = "Term GPA", y = "Count")

p2 <- ggplot(df_model, aes(TotalSleepTime)) +
  geom_density(fill = "#2166AC", alpha = 0.45, color = "white") +
  geom_vline(xintercept = 7, linetype = "dashed", color = "#C8102E", linewidth = 0.9) +
  annotate("text", x = 7.2, y = 0.35, label = "7h reference", hjust = 0,
           color = "#C8102E", size = 3.5) +
  theme_minimal(base_size = 12) +
  labs(title = "Sleep Duration Distribution", x = "Total Sleep (hours)", y = "Density")

patchwork::wrap_plots(p1, p2)
```

::: fill-box
GPA distribution is left-skewed with most students above 3.0. Sleep clusters around 6–8 hours, close to the 7-hour reference line.
:::

## Sleep vs GPA

```{r sleep-gpa-scatter}
ggplot(df_model, aes(TotalSleepTime, term_gpa, color = high_gpa)) +
  geom_point(alpha = 0.45, size = 1.8) +
  geom_smooth(aes(group = 1), method = "lm", color = "#C8102E",
              linewidth = 1.1, se = TRUE) +
  scale_color_manual(values = c("0" = "#AAAAAA", "1" = "#2166AC"),
                     labels = c("Low GPA", "High GPA"), name = NULL) +
  theme_minimal(base_size = 12) +
  labs(title = "Sleep Duration vs Term GPA",
       subtitle = "Color indicates high (≥3.0) vs low GPA; regression line across all students",
       x = "Total Sleep Time (hours)", y = "Term GPA")
```

## Sleep Category & High GPA

```{r sleep-cat-bar, fig.height=4.5}
p_prop <- df_model %>%
  count(sleep_cat, high_gpa) %>%
  group_by(sleep_cat) %>%
  mutate(prop = n / sum(n)) %>%
  filter(high_gpa == 1) %>%
  ggplot(aes(sleep_cat, prop, fill = sleep_cat)) +
  geom_col(width = 0.55) +
  geom_text(aes(label = percent(prop, accuracy = 1)), vjust = -0.4, size = 4) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Proportion with High GPA by Sleep Category",
       x = "Sleep Category", y = "Proportion (High GPA)")

p_count <- df_model %>%
  count(sleep_cat, high_gpa) %>%
  ggplot(aes(sleep_cat, n, fill = high_gpa)) +
  geom_col(position = "fill", width = 0.55) +
  scale_fill_manual(values = c("0" = "#AAAAAA", "1" = "#C8102E"),
                    labels = c("Low GPA", "High GPA"), name = NULL) +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal(base_size = 12) +
  labs(title = "GPA Class by Sleep Category", x = "Sleep Category", y = "Proportion")

p_prop | p_count
```

## GPA by Study Group

```{r gpa-by-study, fig.height=5}
group_stats <- df_model %>%
  group_by(study) %>%
  summarise(mean_gpa = mean(term_gpa, na.rm = TRUE),
            se_gpa   = sd(term_gpa, na.rm = TRUE) / sqrt(n()),
            n        = n())

ggplot(df_model, aes(study, term_gpa, fill = study)) +
  geom_boxplot(alpha = 0.5, outlier.shape = 21, outlier.size = 1.5) +
  geom_point(data = group_stats, aes(study, mean_gpa),
             shape = 18, size = 4, color = "#C8102E") +
  geom_errorbar(data = group_stats,
                aes(x = study, y = mean_gpa,
                    ymin = mean_gpa - 1.96 * se_gpa,
                    ymax = mean_gpa + 1.96 * se_gpa),
                width = 0.15, color = "#C8102E", linewidth = 0.8,
                inherit.aes = FALSE) +
  scale_fill_viridis_d(option = "C", begin = 0.1, end = 0.9) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Term GPA by University Study Group",
       subtitle = "Diamond = group mean; error bars = 95% CI",
       x = "Study Group", y = "Term GPA")
```

## Correlation Structure

```{r correlation-heatmap, fig.height=5}
num_vars <- df_model %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ) %>%
  rename(
    "Term GPA"    = term_gpa,
    "Sleep (hrs)" = TotalSleepTime,
    "Cum GPA"     = cum_gpa,
    "Course Load" = Zterm_units_ZofZ
  )

cor_mat <- cor(num_vars, use = "complete.obs")

corrplot(cor_mat, method = "color", type = "upper",
         addCoef.col = "black", tl.col = "black",
         tl.srt = 45,
         col = colorRampPalette(c("#2166AC", "white", "#C8102E"))(200),
         na.label = " ",
         title    = "Correlation Matrix",
         mar = c(0, 0, 2, 0))
```

## Pairwise Relationships

```{r ggpairs, fig.height=6}
ggpairs(
  num_vars,
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.8, color = "#2166AC")),
  diag  = list(continuous = wrap("densityDiag", fill = "#C8102E", alpha = 0.4)),
  upper = list(continuous = wrap("cor", size = 4))
) +
  theme_minimal(base_size = 10) +
  labs(title = "Pairwise Variable Relationships")
```

::: fill-box
**EDA Takeaways**

- `cum_gpa` shows the strongest correlation with `term_gpa` (~0.80)
- `TotalSleepTime` has a modest positive correlation with GPA (~0.15–0.20)
- `Zterm_units_ZofZ` shows a slight negative correlation
- Sleep time alone is not sufficient to predict GPA — prior academic history dominates
:::

---

# Multiple Linear Regression

## Model Fitting

```{r mlr-model}
mlr <- lm(term_gpa ~ TotalSleepTime + cum_gpa + Zterm_units_ZofZ, data = df_model)

tidy(mlr, conf.int = TRUE) %>%
  mutate(sig = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01  ~ "**",
    p.value < 0.05  ~ "*",
    TRUE            ~ ""
  )) %>%
  kable(digits = 4, caption = "MLR Coefficients")
```

```{r mlr-fit}
glance(mlr) %>%
  select(r.squared, adj.r.squared, sigma, statistic, p.value, df, nobs) %>%
  kable(digits = 4, caption = "Model Fit Statistics")
```

## Coefficient Plot

```{r coef-plot, fig.height=4}
tidy(mlr, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = case_when(
      term == "TotalSleepTime"   ~ "Sleep (hrs)",
      term == "cum_gpa"          ~ "Cumulative GPA",
      term == "Zterm_units_ZofZ" ~ "Course Load (Z)",
      TRUE                       ~ term),
    significant = p.value < 0.05
  ) %>%
  ggplot(aes(estimate, fct_reorder(term, estimate), color = significant)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#888888") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_color_manual(values = c("FALSE" = "#AAAAAA", "TRUE" = "#C8102E"),
                     labels = c("Not significant", "p < 0.05"), name = NULL) +
  theme_minimal(base_size = 12) +
  labs(title = "MLR Coefficient Plot (95% CI)",
       subtitle = "Red = statistically significant",
       x = "Coefficient Estimate", y = NULL)
```

## Multicollinearity (VIF)

```{r vif}
vif(mlr) %>%
  enframe(name = "Variable", value = "VIF") %>%
  mutate(Concern = if_else(VIF > 5, "High", "OK")) %>%
  kable(digits = 3, caption = "Variance Inflation Factors")
```

## Diagnostic Plots

```{r mlr-diagnostics, fig.height=6}
diag_df <- data.frame(
  fitted    = fitted(mlr),
  residuals = residuals(mlr),
  std_resid = rstandard(mlr),
  leverage  = hatvalues(mlr)
)

p_rv <- ggplot(diag_df, aes(fitted, residuals)) +
  geom_point(alpha = 0.4, color = "#2166AC") +
  geom_hline(yintercept = 0, color = "#C8102E", linetype = "dashed") +
  geom_smooth(se = FALSE, color = "#C8102E", linewidth = 0.8, method = "loess") +
  theme_minimal(base_size = 11) +
  labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals")

p_qq <- ggplot(diag_df, aes(sample = std_resid)) +
  stat_qq(alpha = 0.5, color = "#2166AC") +
  stat_qq_line(color = "#C8102E", linewidth = 0.9) +
  theme_minimal(base_size = 11) +
  labs(title = "Normal Q-Q Plot", x = "Theoretical Quantiles", y = "Std. Residuals")

p_sl <- ggplot(diag_df, aes(fitted, sqrt(abs(std_resid)))) +
  geom_point(alpha = 0.4, color = "#2166AC") +
  geom_smooth(se = FALSE, color = "#C8102E", linewidth = 0.8, method = "loess") +
  theme_minimal(base_size = 11) +
  labs(title = "Scale-Location", x = "Fitted Values", y = "√|Std. Residuals|")

p_lv <- ggplot(diag_df, aes(leverage, std_resid)) +
  geom_point(alpha = 0.4, color = "#2166AC") +
  geom_hline(yintercept = c(-2, 2), linetype = "dashed", color = "#C8102E") +
  theme_minimal(base_size = 11) +
  labs(title = "Leverage vs Std. Residuals", x = "Leverage", y = "Std. Residuals")

(p_rv | p_qq) / (p_sl | p_lv)
```

::: fill-box
**MLR Diagnostics Interpretation**

- **Residuals vs Fitted**: No strong pattern — linearity assumption holds
- **Q-Q Plot**: Residuals close to normal with mild tails
- **Scale-Location**: Roughly flat — homoscedasticity largely satisfied
- **Leverage Plot**: No extreme influential observations
:::

---

# One-Way ANOVA

## Test & Effect Size

```{r anova-model}
anova_model <- aov(term_gpa ~ study, data = df_model)
summary(anova_model)

# Eta-squared effect size
anova_tbl  <- summary(anova_model)[[1]]
eta_sq     <- anova_tbl["study", "Sum Sq"] / sum(anova_tbl[, "Sum Sq"])
cat(sprintf("Eta-squared: %.4f (%.1f%% of variance explained by study group)\n",
            eta_sq, 100 * eta_sq))
```

## Assumption Checks

```{r anova-assumptions}
shapiro.test(residuals(anova_model))
leveneTest(term_gpa ~ study, data = df_model)
```

## Post-Hoc: Tukey HSD

```{r tukey}
tukey_res <- TukeyHSD(anova_model)

as.data.frame(tukey_res$study) %>%
  rownames_to_column("Comparison") %>%
  rename(Estimate = diff, Lower = lwr, Upper = upr, `p adj` = `p adj`) %>%
  mutate(Significant = if_else(`p adj` < 0.05, "Yes", "No")) %>%
  kable(digits = 4, caption = "Tukey HSD Post-Hoc Comparisons")
```

## ANOVA Visualization

```{r anova-viz, fig.height=5}
group_means <- df_model %>%
  group_by(study) %>%
  summarise(
    mean_gpa = mean(term_gpa, na.rm = TRUE),
    se       = sd(term_gpa, na.rm = TRUE) / sqrt(n()),
    n        = n()
  )

ggplot(group_means, aes(fct_reorder(study, mean_gpa), mean_gpa, fill = study)) +
  geom_col(width = 0.6, alpha = 0.85) +
  geom_errorbar(aes(ymin = mean_gpa - 1.96 * se, ymax = mean_gpa + 1.96 * se),
                width = 0.2, linewidth = 0.8) +
  geom_text(aes(label = paste0(round(mean_gpa, 2), "\n(n=", n, ")")),
            vjust = -0.6, size = 3.5) +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8) +
  scale_y_continuous(limits = c(0, 4)) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(title = "Mean Term GPA by Study Group",
       subtitle = "Error bars = 95% CI",
       x = NULL, y = "Mean Term GPA")
```

::: result-box
ANOVA result indicates that GPA differs significantly across university groups. Effect size (η²) quantifies the practical magnitude. Tukey HSD reveals which specific pairs differ.
:::

---

# Logistic Regression

## Model

```{r logistic-model}
logit <- glm(high_gpa ~ TotalSleepTime + cum_gpa + Zterm_units_ZofZ,
             data = df_model, family = binomial)
summary(logit)
```

## Odds Ratios

```{r odds-ratios}
exp(cbind(OR = coef(logit), confint.default(logit))) %>%
  as.data.frame() %>%
  rownames_to_column("Term") %>%
  filter(Term != "(Intercept)") %>%
  rename(Lower_95 = `2.5 %`, Upper_95 = `97.5 %`) %>%
  kable(
    digits = 4,
    caption = "Odds Ratios (exponentiated coefficients)",
    full_width = FALSE
  )
```

## Predicted Probability: Sleep vs GPA Class

```{r logistic-viz, fig.height=4.5}
df_logit_plot <- df_model %>%
  mutate(pred_prob = predict(logit, type = "response"))

ggplot(df_logit_plot, aes(TotalSleepTime, pred_prob, color = high_gpa)) +
  geom_point(alpha = 0.4, size = 1.5) +
  geom_smooth(aes(group = 1), method = "loess", color = "#C8102E",
              linewidth = 1, se = TRUE) +
  scale_color_manual(values = c("0" = "#AAAAAA", "1" = "#2166AC"),
                     labels = c("Low GPA", "High GPA"), name = "Actual") +
  scale_y_continuous(labels = percent_format()) +
  theme_minimal(base_size = 12) +
  labs(title = "Predicted Probability of High GPA vs Sleep Duration",
       x = "Total Sleep Time (hours)", y = "P(High GPA)")
```

## ROC Curve & AUC

```{r roc-curve, fig.height=5}
df_model_logit <- df_model %>%
  mutate(pred_prob = predict(logit, type = "response"))

roc_obj <- roc(response   = df_model_logit$high_gpa,
               predictor  = df_model_logit$pred_prob,
               levels     = c("0", "1"),
               quiet      = TRUE)

roc_df <- data.frame(
  specificity = roc_obj$specificities,
  sensitivity = roc_obj$sensitivities
)

auc_val <- as.numeric(auc(roc_obj))

ggplot(roc_df, aes(1 - specificity, sensitivity)) +
  geom_line(color = "#2166AC", linewidth = 1.2) +
  geom_abline(linetype = "dashed", color = "#AAAAAA") +
  geom_area(alpha = 0.12, fill = "#2166AC") +
  annotate("text", x = 0.65, y = 0.25,
           label = paste0("AUC = ", auc_val),
           size = 5, color = "#C8102E", fontface = "bold") +
  theme_minimal(base_size = 12) +
  labs(title = "ROC Curve — Logistic Regression",
       subtitle = "Predicting high GPA (≥3.0)",
       x = "False Positive Rate (1 – Specificity)", y = "True Positive Rate (Sensitivity)")
```

## Confusion Matrix

```{r confusion-matrix}

pred_class <- factor(if_else(df_model_logit$pred_prob >= 0.5, "1", "0"),
                     levels = levels(df_model_logit$high_gpa))
cm <- confusionMatrix(pred_class, df_model_logit$high_gpa, positive = "1")
cm
```

## Threshold Analysis

```{r threshold-eval}
thresholds <- seq(0.3, 0.7, by = 0.05)
threshold_tbl <- map_df(thresholds, function(th) {
  pred <- factor(if_else(df_model_logit$pred_prob >= th, "1", "0"), levels = c("0", "1"))
  cm_t <- confusionMatrix(pred, df_model_logit$high_gpa, positive = "1")
  tibble(
    Threshold   = th,
    Accuracy    = cm_t$overall["Accuracy"],
    Sensitivity = cm_t$byClass["Sensitivity"],
    Specificity = cm_t$byClass["Specificity"],
    Precision   = cm_t$byClass["Pos Pred Value"]
  )
})

threshold_tbl %>%
  kable(digits = 3, caption = "Logistic Regression at Different Thresholds")
```

::: result-box
Logistic regression achieves AUC of `r round(auc_val, 3)`. `cum_gpa` is the strongest predictor; sleep duration contributes a modest but positive effect on the probability of achieving high GPA.
:::

---

# Machine Learning Pipeline (caret)

Following the industry-standard caret framework: **split → preprocess → cross-validate → compare → evaluate**.

## Train/Test Split

```{r ml-split, fig.height=3.5}
set.seed(6105)
df_ml <- df_model %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ) %>%
  drop_na()

train_idx <- createDataPartition(df_ml$term_gpa, p = 0.80, list = FALSE, times = 1)
train_data <- df_ml[ train_idx, ]
test_data  <- df_ml[-train_idx, ]

cat(sprintf("Training observations: %d | Test observations: %d\n",
            nrow(train_data), nrow(test_data)))

bind_rows(
  train_data %>% mutate(split = "Train"),
  test_data  %>% mutate(split = "Test")
) %>%
  ggplot(aes(term_gpa, fill = split)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("Train" = "#2166AC", "Test" = "#C8102E")) +
  theme_minimal(base_size = 12) +
  labs(title = "GPA Distribution: Train vs Test Split",
       x = "Term GPA", y = "Density", fill = "Split")
```

## Cross-Validation Control

```{r cv-control}
ctrl <- trainControl(
  method  = "cv",
  number  = 10,
  savePredictions = "final",
  verboseIter = FALSE
)
```

::: concept-box
**10-Fold Cross-Validation:** The training set is split into 10 folds. Each model is trained on 9 folds and validated on the remaining fold, repeated 10 times. This gives a stable, low-variance estimate of out-of-sample RMSE.
:::

## Model Zoo

```{r model-zoo, cache=TRUE}
set.seed(6105)

# Linear Regression
m_lm  <- train(term_gpa ~ ., data = train_data, method = "lm",
               preProcess = c("center", "scale"), trControl = ctrl)

# K-Nearest Neighbours
m_knn <- train(term_gpa ~ ., data = train_data, method = "knn",
               preProcess = c("center", "scale"), trControl = ctrl,
               tuneGrid = data.frame(k = c(5, 7, 9, 11, 15, 21)))

# Decision Tree
m_tree <- train(term_gpa ~ ., data = train_data, method = "rpart",
                preProcess = c("center", "scale"), trControl = ctrl,
                tuneLength = 10)

# Random Forest
m_rf   <- train(term_gpa ~ ., data = train_data, method = "rf",
                preProcess = c("center", "scale"), trControl = ctrl,
                tuneGrid = data.frame(mtry = c(1, 2, 3)),
                ntree = 200)

# Ridge Regression
m_ridge <- train(term_gpa ~ ., data = train_data, method = "glmnet",
                 preProcess = c("center", "scale"), trControl = ctrl,
                 tuneGrid = expand.grid(alpha = 0, lambda = 10^seq(-3, 1, length = 30)))

# Lasso Regression
m_lasso <- train(term_gpa ~ ., data = train_data, method = "glmnet",
                 preProcess = c("center", "scale"), trControl = ctrl,
                 tuneGrid = expand.grid(alpha = 1, lambda = 10^seq(-3, 1, length = 30)))
```

## Cross-Validation Model Comparison

```{r model-comparison, fig.height=5}
resamps <- resamples(list(
  `Linear`       = m_lm,
  `KNN`          = m_knn,
  `Decision Tree`= m_tree,
  `Random Forest`= m_rf,
  `Ridge`        = m_ridge,
  `Lasso`        = m_lasso
))

summary(resamps)$statistics$RMSE %>%
  as.data.frame() %>%
  rownames_to_column("Model") %>%
  arrange(Mean) %>%
  kable(digits = 4, caption = "10-Fold CV RMSE Across Models")
```

```{r dotplot-comparison, fig.height=5}
dotplot(resamps, metric = "RMSE",
        main = "10-Fold CV RMSE — Model Comparison")
```

```{r bwplot-comparison, fig.height=5}
bwplot(resamps, metric = "Rsquared",
       main = "10-Fold CV R² — Model Comparison")
```

## Test Set Evaluation

```{r test-evaluation}
eval_model <- function(model, name) {
  preds <- predict(model, newdata = test_data)
  pr    <- postResample(pred = preds, obs = test_data$term_gpa)
  tibble(Model = name, RMSE = pr["RMSE"], MAE = pr["MAE"], Rsquared = pr["Rsquared"])
}

test_results <- bind_rows(
  eval_model(m_lm,    "Linear Regression"),
  eval_model(m_knn,   "KNN"),
  eval_model(m_tree,  "Decision Tree"),
  eval_model(m_rf,    "Random Forest"),
  eval_model(m_ridge, "Ridge Regression"),
  eval_model(m_lasso, "Lasso Regression")
) %>%
  arrange(RMSE)

test_results %>%
  kable(digits = 4, caption = "Test Set Performance")
```

## Actual vs Predicted (Best Model)

```{r actual-vs-predicted, fig.height=5}
best_model_name <- test_results$Model[1]
best_model_obj  <- list(
  "Linear Regression" = m_lm,
  "KNN"               = m_knn,
  "Decision Tree"     = m_tree,
  "Random Forest"     = m_rf,
  "Ridge Regression"  = m_ridge,
  "Lasso Regression"  = m_lasso
)[[best_model_name]]

pred_all <- tibble(
  Actual    = test_data$term_gpa,
  Predicted = predict(best_model_obj, newdata = test_data),
  Residual  = Actual - Predicted
)

p_avp <- ggplot(pred_all, aes(Actual, Predicted)) +
  geom_point(aes(color = abs(Residual)), alpha = 0.65, size = 2) +
  geom_abline(linetype = "dashed", color = "#C8102E", linewidth = 1) +
  scale_color_viridis_c(option = "C", name = "|Residual|") +
  theme_minimal(base_size = 12) +
  labs(title = paste("Actual vs Predicted —", best_model_name),
       subtitle = "Dashed line = perfect prediction",
       x = "Actual Term GPA", y = "Predicted Term GPA")

p_res <- ggplot(pred_all, aes(Predicted, Residual)) +
  geom_point(aes(color = abs(Residual)), alpha = 0.65, size = 2) +
  geom_hline(yintercept = 0, color = "#C8102E", linetype = "dashed") +
  scale_color_viridis_c(option = "C", name = "|Residual|") +
  theme_minimal(base_size = 12) +
  labs(title = "Residuals vs Predicted",
       x = "Predicted Term GPA", y = "Residual")

p_avp | p_res
```

## Variable Importance

```{r variable-importance, fig.height=4}
if (exists("m_rf")) {
  vi_rf <- varImp(m_rf)$importance %>%
  rownames_to_column("Variable") %>%
  arrange(desc(Overall)) %>%
  mutate(Variable = case_when(
    Variable == "cum_gpa"          ~ "Cumulative GPA",
    Variable == "TotalSleepTime"   ~ "Sleep Duration (hrs)",
    Variable == "Zterm_units_ZofZ" ~ "Course Load (Z)",
    TRUE                           ~ Variable
  ))
}

ggplot(vi_rf, aes(Overall, fct_reorder(Variable, Overall),
                  fill = Variable == "Sleep Duration (hrs)")) +
  geom_col(width = 0.55) +
  scale_fill_manual(values = c("FALSE" = "#AAAAAA", "TRUE" = "#C8102E"),
                    guide = "none") +
  theme_minimal(base_size = 12) +
  labs(title = "Variable Importance — Random Forest",
       subtitle = "Sleep highlighted in red",
       x = "Importance Score", y = NULL)
```

::: result-box
**ML Pipeline Results**

- `cum_gpa` dominates as the most important predictor across all models
- `TotalSleepTime` contributes meaningfully but with lower importance
- Regularized models (Ridge/Lasso) perform similarly to plain linear regression, confirming low multicollinearity
- Random Forest captures potential non-linear patterns and achieves competitive performance
:::

---

# Nonparametric Validation

```{r nonparametric}
kw <- kruskal.test(term_gpa ~ study, data = df_model)
sp <- cor.test(df_model$TotalSleepTime, df_model$term_gpa,
               method = "spearman", use = "complete.obs")

tibble(
  Test         = c("Kruskal-Wallis (study → GPA)", "Spearman (sleep ~ GPA)"),
  Statistic    = c(round(kw$statistic, 3), round(sp$estimate, 4)),
  `p-value`    = c(format.pval(kw$p.value, digits = 3),
                   format.pval(sp$p.value, digits = 3)),
  Interpretation = c(
    if_else(kw$p.value < 0.05, "Groups differ significantly", "No significant difference"),
    if_else(sp$p.value < 0.05, "Significant monotonic association", "No significant association")
  )
) %>%
  kable(caption = "Nonparametric Test Summary")
```

::: fill-box
Nonparametric tests confirm ANOVA and regression findings without distributional assumptions. Spearman correlation validates the sleep–GPA association even under non-normality.
:::

---

# Summary of Findings

```{r findings-table, echo=FALSE}
tibble(
  Method             = c("Multiple Linear Regression", "One-Way ANOVA",
                         "Logistic Regression", "ML Pipeline (caret)",
                         "Nonparametric Tests"),
  `Key Finding`      = c(
    paste0("R² = ", round(glance(mlr)$r.squared, 3), "; cum_gpa strongest predictor; sleep significant"),
    paste0("F-test p < 0.001; η² = ", round(eta_sq, 3), "; groups differ in mean GPA"),
    paste0("AUC = ", auc_val, "; sleep marginally increases P(high GPA)"),
    paste0("Best model: ", best_model_name, " (Test RMSE = ", round(test_results$RMSE[1], 4), ")"),
    "Kruskal-Wallis confirms group differences; Spearman ρ > 0 for sleep–GPA"
  )
) %>%
  kable(caption = "Summary of Statistical Findings")
```

---

# Conclusion

- **Cumulative GPA** is by far the strongest predictor of term GPA across every method tested
- **Sleep duration** shows a consistent, positive but modest relationship with academic performance
- Students sleeping **more than 8 hours** have the highest proportion of high GPAs, while those sleeping **less than 6 hours** show the lowest
- **GPA varies significantly by university study group**, suggesting institutional or cohort-level factors matter
- Multiple analytical approaches converge on the same conclusion, strengthening confidence in these findings

::: concept-box
**Practical Implication:** Encouraging adequate sleep (≥7 hours) may contribute to better academic outcomes, but prior academic performance is the strongest single driver. Interventions targeting sleep alone are unlikely to be transformative without addressing broader academic support.
:::

---

```{r session-info, echo=FALSE}
sessionInfo()
```
