---
title: "INFO 6105 Final Project"
subtitle: "Impact of Sleep Duration on Academic Performance (GPA)"
author: "Rushita Vachhani"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)

library(tidyverse)
library(broom)
library(car)
library(pROC)
library(caret)
library(here)
library(GGally)
```

```{=html}
<style>
body { font-family: 'Palatino', 'Georgia', serif; }
h1, h2 { color: #C8102E; border-bottom: 2px solid #C8102E; padding-bottom: 4px; }
h3 { color: #444444; }
.fill-box {
  border: 2px solid #C8102E; padding: 12px; margin: 10px 0;
  background-color: #fdf5f5; border-radius: 5px;
}
.concept-box {
  border: 2px solid #2166AC; padding: 12px; margin: 10px 0;
  background-color: #f0f6fc; border-radius: 5px;
}
</style>
```
------------------------------------------------------------------------


# Introduction

This project investigates whether **sleep duration impacts academic performance (GPA)** using real-world student data.

We apply multiple statistical methods to evaluate this relationship while controlling for additional academic factors.


```{r}
df <- read.csv(here("data/raw/cmu-sleep.csv"))

df_clean <- df %>%
  select(term_gpa, TotalSleepTime, cum_gpa, Zterm_units_ZofZ, study) %>%
  mutate(
    TotalSleepTime = TotalSleepTime / 60,
    high_gpa = if_else(term_gpa >= 3.0, 1, 0),
    high_gpa = factor(high_gpa, levels = c(0, 1)),
    study = factor(study)
  )

glimpse(df_clean)
```

::: fill-box
Key Preparation Steps:
	•	Converted sleep time into hours
	•	Created binary outcome (high_gpa)
	•	Converted categorical variables to factors
:::

------------------------------------------------------------------------

# Exploratory Data Analysis

## GPA Distribution

```{r}

ggplot(df_clean, aes(term_gpa)) +
  geom_histogram(fill = "steelblue", bins = 30) +
  labs(title = "Distribution of GPA", x = "GPA", y = "Count")
```

::: fill-box
GPA is approximately normally distributed with concentration in mid-to-high ranges.
:::

## Sleep Distribution

```{r}
ggplot(df_clean, aes(TotalSleepTime)) +
  geom_density(fill = "purple", alpha = 0.4) +
  labs(title = "Sleep Duration Distribution", x = "Hours", y = "Density")

```

## Sleep vs GPA

```{r}
ggplot(df_clean, aes(TotalSleepTime, term_gpa)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Sleep vs GPA", x = "Sleep (hours)", y = "GPA")
```

::: fill-box
A slight positive trend suggests that higher sleep duration may be associated with higher GPA.
:::

## GPA by Study Type

```{r}

ggplot(df_clean, aes(study, term_gpa)) +
  geom_boxplot(fill = "orange") +
  labs(title = "GPA by Study Type", x = "Study", y = "GPA")

```

## Correlation Matrix

```{r}

df_clean %>%
  select(term_gpa, TotalSleepTime, cum_gpa) %>%
  GGally::ggpairs()

```

::: fill-box
Cumulative GPA shows the strongest relationship with term GPA, while sleep has a weaker but positive association.
:::

## Multiple Linear Regression

```{r}

mlr <- lm(term_gpa ~ TotalSleepTime + cum_gpa + Zterm_units_ZofZ, data = df_clean)
summary(mlr)

```

::: fill-box
	•	cum_gpa is expected to be the strongest predictor
	•	Sleep duration may have a smaller positive effect
:::

## Diagnostics

```{r}

par(mfrow = c(2,2))
plot(mlr)

```

## One-Way ANOVA

```{r}

anova_model <- aov(term_gpa ~ study, data = df_clean)
summary(anova_model)

TukeyHSD(anova_model)

```

::: fill-box
ANOVA tests whether GPA differs significantly across study groups.
:::

## Logistic Regression

```{r}
log_model <- glm(high_gpa ~ TotalSleepTime + cum_gpa + Zterm_units_ZofZ,
                 data = df_clean, family = binomial)

summary(log_model)

```


## ROC Curve

```{r}

# Get data used in model (after NA removal)
model_data <- model.frame(log_model)

# Predicted probabilities (same length as model_data)
prob <- predict(log_model, type = "response")

# ROC using aligned data
roc_obj <- roc(model_data$high_gpa, prob)

plot(roc_obj, col = "blue")
auc(roc_obj)

```

## Confusion Matrix

```{r}

# Use same model data
actual <- model_data$high_gpa

pred <- if_else(prob > 0.5, 1, 0) %>%
  factor(levels = c(0,1))

confusionMatrix(pred, actual)
```

::: fill-box
Logistic regression evaluates classification performance of predicting high GPA.
:::

## Nonparametric Analysis

```{r}

kruskal.test(term_gpa ~ study, data = df_clean)

cor.test(df_clean$TotalSleepTime, df_clean$term_gpa, method = "spearman")

```

::: fill-box
Nonparametric tests validate results without strict distributional assumptions.
:::

# Conclusion
	•	Cumulative GPA is the strongest predictor of academic performance
	•	Sleep duration shows a positive but moderate relationship with GPA
	•	Results are consistent across multiple statistical approaches
Overall, maintaining healthy sleep patterns may contribute to improved academic outcomes.








